---
title: "Batch Normalise"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batch Normalise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Create Omiprep object

```{r setup}
library(omiprep)

# example file
filepath <- system.file("extdata", "metabolon_v1.1_example.xlsx", package = "omiprep")

## Read in Metabolon v1.1 example and directly create Omiprep object
mydata <- read_metabolon(filepath, sheet = "OrigScale" , return_Omiprep = TRUE )


## summary of Omiprep object
summary(mydata)
```

## Run batch normalisation

Early Metabolon releases only provided the original scale data and a batch normalised and minimally imputed data set. We would not advocate using minimally imputed data, particularly not for quality control / pre-processing. As such, you may have to (1) batch normalize the original scale data, as we are doing here. Or (2) convert the minimally imputed cells in the ScaledImp tab of the release back into NAs. 

Again, here we are illustrating how to run a mass-spectroscopy style median batch normalization. 

```{r batch_normalise}
mydata <- mydata |>
  batch_normalise(run_mode_col          = "platform",  ## Column names in feature sheet that defines the run mode for each feature
                  run_mode_colmap       = 
                    c("LC/MS Neg"       = 'lcms_neg' , ## string from feature sheet column = 
                                                       ## column name in sample sheet that holds the batch information
                      "LC/MS Polar"     = 'lcms_polar',
                      "LC/MS Pos Early" = 'lcms_pos_early',
                      "LC/MS Pos Late"  = 'lcms_pos_late' ) 
                  )

```


## Accessing data

### Raw input data

```{r data}
mydata@data[1:5, 1:5, "input"]
```

### Batch normalised data

```{r}
mydata@data[1:5, 1:5, "batch_normalised"]
```

## Now any quality control step should be run on the "batch_normalised" layer. 

```{r}
## identify the xenos
xenos <- mydata@features[!is.na(mydata@features$super_pathway) & mydata@features$super_pathway == "Xenobiotics", "feature_id"]

## run the QC on the batch_normalised layer
mydata <- mydata |> quality_control(source_layer               = "batch_normalised", 
                                    sample_missingness         = 0.2, 
                                    feature_missingness        = 0.2, 
                                    total_sum_abundance_sd     = 5, 
                                    outlier_udist              = 5, 
                                    outlier_treatment          = "leave_be", 
                                    winsorize_quantile         = 1.0, 
                                    tree_cut_height            = 0.5, 
                                    pc_outlier_sd              = 5,
                                    feature_selection          = "least_missingness", 
                                    features_exclude_but_keep  = xenos, 
                                    cores                      = 1
                                    )
```

### QC data

```{r}
mydata@data[1:5, 1:5, "qc"]
```

## Quick view of omiprep object

### data array layers

```{r}
dimnames(mydata@data)[[3]]
```

### omiprep summary

```{r}
summary(mydata)
```

