---
title: "Feature summary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feature summary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Create Omiprep object

```{r setup}
library(omiprep)

# import data
data     <- read.csv(system.file("extdata", "dummy_data.csv", package = "omiprep"), header=T, row.names = 1, check.names = FALSE) |> as.matrix()
samples  <- read.csv(system.file("extdata", "dummy_samples.csv", package = "omiprep"), header=T, row.names = 1)
features <- read.csv(system.file("extdata", "dummy_features.csv", package = "omiprep"), header=T, row.names = 1)
features$feature_id = as.character(features$feature_id)

# create object
mydata <- Omiprep(data = data, samples = samples, features = features)

```

## Summary of Omiprep object

```{r omiprep_summary}
summary(mydata)
```

## Run standard quality control

```{r run_qc}
mydata = quality_control(mydata)
```

## Feature Summary

### View feature summary from the QC pipeline

```{r qc_sample_sum_table, echo=TRUE}
# Note: the quality_control() ultimately returns the feature_summary attribute as a matrix.
df <- t( as.data.frame(mydata@feature_summary[, 1:5, "input"]) )
df <- as.data.frame( round( df , 3) )
df <- cbind(feature_id = rownames(df), df)

df |> knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = TRUE) 
```

### Manually run feature summary

While feature summary is run as a part of the quality_control() function pipeline you can run the function yourself, on any layer you wish. 


```{r feature_summary1}
# NOTE:
# outlier_udist = number of IQRs from the median at which a value is flagged.
# 1.0 here is illustrative; in practice we favour 5.0, which is the default value
# for the quality_control() function.
feature_sum1 <- feature_summary(omiprep         = mydata, 
                                source_layer    = "input", 
                                outlier_udist   = 1.0,
                                tree_cut_height = 0.5,
                                output          = "data.frame", 
                                cores           = 1)
```

#### Table of feature summary

```{r feature_sum1_table, echo=TRUE}
feature_sum1 |> 
  head(n = 10) |> 
  knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = FALSE) 
```


### Run feature summary on subset

Using the `sample_ids` and `feature_ids` arguments you can run the summary for a subset of the data. Note: all rows will be return, however summary data will only be returned for the specified ids.  

```{r feature_sum_subset}
## define a vector of sample IDs
sids <- mydata@samples[mydata@samples$sex == "female", "sample_id"] 

## define a vector of feature IDs
fids <- mydata@features[, "feature_id"] |> sample(25)

# NOTE:
# outlier_udist = number of IQRs from the median at which a value is flagged.
# 1.0 here is illustrative; in practice we favour 5.0, which is the default value
# for the quality_control() function.
feature_sum_subset <- feature_summary(omiprep         = mydata, 
                                      source_layer    = "input", 
                                      outlier_udist   = 1.0,
                                      tree_cut_height = 0.5,
                                      sample_ids      = sids,
                                      feature_ids     = fids,
                                      output          = "data.frame",
                                      cores           = 1)
```


#### Table of feature summary for subset

```{r feature_sum_subset_table, echo=TRUE}
feature_sum_subset |> 
  na.omit() |>
  knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::scroll_box(width = "100%", height = "500px")
```


## Additional feature_summary() attributes

```{r}
## The attributes include column names, row names, and class for the feature summary table
## as well as a hierarchical cluster dendrogram or `input_tree` and the parameter values for 
## outlier_udist and input_tree_cut_height passed to the function. 
names( attributes(feature_sum1) )
```

### hierarchical cluster dendrogram

In addition to the summary data, the hierarchical cluster dendrogram is appended to the returned `data.frame` as and `attribute`. This can be accessed with the attribute name: `[source_layer]_tree`, in this case we summarised the `input` data, therefore the attribute name is `input_tree`. 


```{r tree, out.width="100%", fig.align='center', fig.alt = "Decision tree showing feature importance in dataset", fig.width = 20, fig.height = 5}
suppressPackageStartupMessages(library(dendextend))

## number of independent features
indfeatcount = sum( feature_sum1$independent_features, na.rm = TRUE )

# extract tree from attributes
tree <- attr(feature_sum1, 'input_tree')
dend <- stats::as.dendrogram(tree)

# color the independent features blue
metab_color       <- feature_sum1[, c("feature_id", "independent_features")]
metab_color       <- metab_color[match(labels(dend), metab_color$feature_id), ]
metab_color$color <- ifelse(metab_color$independent_features==TRUE, "#084594", "grey80")

# format dendrogram for ploting
dend <- dend |>
  dendextend::set("labels_cex", 0.75) |>
  dendextend::set("labels_col", metab_color$color) |>
  dendextend::set("branches_lwd", 1) |>
  dendextend::set("branches_k_color",  value = metab_color$color)

## plot the dendrogram
dend |> plot(main  = paste0("Feature clustering dendrogram\n# of ind. features = ",indfeatcount ))
abline(h = 0.5, col = "#E41A1C", lwd = 1.5)

```


## Run sample & feature summaries together

```{r summarise_data}
# NOTE:
# outlier_udist = number of IQRs from the median at which a value is flagged.
# 1.0 here is illustrative; in practice we favour 5.0, which is the default value
# for the quality_control() function.
sf_sum <- summarise(omiprep         = mydata, 
                    source_layer    = "input", 
                    outlier_udist   = 1.0,
                    tree_cut_height = 0.5,
                    sample_ids      = sids, ## It is also possible to run on a subset of samples and/or features
                    feature_ids     = fids,
                    output          = "data.frame", 
                    cores           = 1)

```

## Table of feature summary from summarise() function

```{r show2, echo=TRUE}
sf_sum$feature_summary |> 
  na.omit() |>
  knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::scroll_box(width = "100%", height = "500px")
```