---
title: "Import Metabolon Metabolomic Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Import Metabolon Metabolomic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Import Metabolon data

Read in the Metabolon data using the `read_metabolon` function. Here we will read in the example data provided with the package, as a list object.

```{r setup}
library(omiprep)

# example file
filepath <- system.file("extdata", "metabolon_v2_example.xlsx", package = "omiprep")

# import data as a list object rather than directly as a Omiprep object
datain <- read_metabolon(filepath,  
                         sheet = 'Batch-normalized Data',        ## read in the batch normalized but NOT imputed tab
                         feature_sheet = "Chemical Annotation",  ## tab name of the feature data
                         feature_id_col = "COMP_ID",             ## in this instance features in batch normalized.. 
                                                                 ##  ..tab are annotated to column name COMP_ID. Verify in your data.
                         sample_sheet = "Sample Meta Data",      ## tab name of the sample data
                         sample_id_col = "PARENT_SAMPLE_NAME",   ## column name of sample IDs 
                         return_Omiprep = FALSE                  ## return data as a list of data, feature, and samples, ..
                                                                 ##  .. NOT an omiprep object  
                         )
```

## Quick look at data structure of the imported data

```{r data_str}
str(datain)
```


## Create Omiprep object 

Once imported, we pass the data to the Omiprep() function to build the `Omiprep` class object. 

```{r omioprep_obj}
## This step could be avoided by defining return_Omiprep = TRUE in read_metabolon() function above.
mydata <- Omiprep(data     = datain$data, 
                  features = datain$features, 
                  samples  = datain$samples)
```


## Quick summary of the Omiprep object

```{r summary_before_qc}
summary(mydata)
```


## Identify the Xenobiotics to exclude from the QC steps

Use the feature data just imported to identify xenobiotic metabolites.  It may be best to excluded these features from the quality-control (QC) process. Xenobiotics typically exhibit much higher levels of missingness than endogenous metabolites, and including them in QC can result in excessive exclusion of both features and samples. This step will allow you to retain these features in the final dataset, by excluding them from QC filtering steps.


```{r xeno_identification}
xenos <- mydata@features[!is.na(mydata@features$SUPER_PATHWAY) & 
                           mydata@features$SUPER_PATHWAY == "Xenobiotics", "feature_id"]

## how many xenobiotics identified
length(xenos)
```

## QC Metabolon

Perform the QC steps using the `quality_control` function, specifying the xenobiotics to exclude from the QC steps.

```{r qc}
## Given the high missingness in metabolon data, 
##   we suggest using the `least_missingness` feature selection method
##   for the identification of principle variable that will then be 
##   used in the construction of PCs. 

mydata <- mydata |>
  quality_control(source_layer              = "input", 
                  sample_missingness        = 0.2, 
                  feature_missingness       = 0.2, 
                  total_sum_abundance_sd    = 5, 
                  outlier_udist             = 5, 
                  outlier_treatment         = "leave_be", 
                  winsorize_quantile        = 1.0, 
                  tree_cut_height           = 0.5, 
                  pc_outlier_sd             = 5,
                  feature_selection         = "least_missingness", ## We suggest using `least_missingness` 
                                                                   ## when working with data, like Metabolon, 
                                                                   ## with high missingness. 
                                                                   ## Default is "max_var_exp".
                  features_exclude_but_keep = xenos, ## Exclude xenobiotics from QC, but retain them in the final dataset.
                                                     ## We suggest this as xenos can have extreme missingness and are 
                                                     ## commonly qc'd from a data set. However, they may be appropriate 
                                                     ## to model as a binary present|absent trait. A choice for the researcher.
                  cores                     = 1
                  )

```

## Quick summary of the Omiprep object following QC

```{r summary_after_qc}
summary(mydata)
```