---
title: "Sample summary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sample summary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Create Omiprep object

```{r setup}
library(omiprep)

# import data
data     <- read.csv(system.file("extdata", "dummy_data.csv", package = "omiprep"), header=T, row.names = 1, check.names = FALSE) |> as.matrix()
samples  <- read.csv(system.file("extdata", "dummy_samples.csv", package = "omiprep"), header=T, row.names = 1)
features <- read.csv(system.file("extdata", "dummy_features.csv", package = "omiprep"), header=T, row.names = 1)
features$feature_id = as.character(features$feature_id)

# create object
mydata <- Omiprep(data = data, samples = samples, features = features)
```

## Summary of Omiprep object

```{r omiprep_summary}
summary(mydata)
```

## Run standard quality control

```{r run_qc}
mydata = quality_control(mydata)
```

## Sample Summary

### View sample summary from the QC pipeline

```{r qc_sample_sum_table, echo=TRUE}
# Note: the quality_control() ultimately returns the sample_summary attribute as a matrix.
df <- as.data.frame(mydata@sample_summary[1:3, 1:6, "input"])
df <- cbind(sample_id = rownames(df), df)

df |> knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = TRUE) 
```


### Manually run sample summary

While sample summary is run as a part of the quality_control() function pipeline you can run the function yourself, on any layer you wish. 

```{r sample_summary1}
# NOTE:
# outlier_udist = number of IQRs from the median at which a value is flagged.
# 1.0 here is illustrative; in practice we favour 5.0, which is the default value
# for the quality_control() function.
sample_sum1 <- sample_summary(omiprep         = mydata, 
                              source_layer    = "input", 
                              outlier_udist   = 1.0,
                              output          = "data.frame")
```

#### Table of sample summary

```{r sample_sum1_table, echo=TRUE}
sample_sum1 |> 
  head(n = 10) |> 
  knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = TRUE) 
```


### Run sample summary on subset

Using the `sample_ids` and `feature_ids` arguments you can run the summary for a subset of the data. Note: all rows will be return, however summary data will only be returned for the specified ids.  

```{r sample_summary_subset}
## define a vector of sample IDs
sids <- mydata@samples[mydata@samples$sex == "female", "sample_id"] 
cat("Number of samples being passed = ", length(sids), "\n")

## define a vector of feature IDs
## extract only those features run on `pos`itive ion mode. 
fids <- mydata@features[mydata@features$platform == "LC/MS Pos Late", "feature_id"] 
cat("Number of features being passed = ", length(fids), "\n")

# run sample summary on subset
sample_sum_subset <- sample_summary(omiprep       = mydata, 
                                    source_layer  = "input", 
                                    outlier_udist = 1.0,
                                    sample_ids    = sids,
                                    feature_ids   = fids,
                                    output        = "data.frame")
```

#### Table of sample summary on subset

```{r sample_sum_subset_table, echo=TRUE}
sample_sum_subset |> 
  na.omit() |>
  head(n = 10) |> 
  knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = TRUE) 
```


## Principal Componet Analysis

### View PCs from the QC pipeline

PCs and outliers are available as a part of the `quality_countrol()` function pipeline.

```{r qc_PCs_table, echo=TRUE}
# Note: the quality_control() ultimately returns the sample_summary attribute as a matrix.
df <- as.data.frame(mydata@sample_summary[1:3, -c(1:6), "input"])
df <- cbind(sample_id = rownames(df), df)

df |> knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = TRUE) 
```

### Manually run PCA analysis

You can derive PCs and identify outlier independent of the `quality_control()` function, however.

`pc_and_outliers()` performs principal component analysis. Missing data is imputed to the median and used to identify the number of informative or 'significant' PCs by (1) an acceleration analysis, and (2) a parallel analysis. Finally the number of sample outliers are determined at 3, 4, and 5 standard deviations from the mean on the top PCs as determined by the acceleration factor analysis.

```{r PC_analysis}
pc_analysis <- pc_and_outliers(omiprep      = mydata, 
                               source_layer = "input",
                               sample_ids   = sids, ## It is also possible to run on a subset of samples and/or features
                               feature_ids  = NULL
                               )
```


#### Table of PCA analysis results

Returned are the PC eigenvectors for the top 10 PCs, and outlier counts at 3, 4, and 5 standard deviations from the mean for the top two PCs

```{r PC_table, echo=TRUE}
pc_analysis |> 
  head(n = 10) |> 
  knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = TRUE) 
```

#### PCA plot

```{r pca_plot}
library(ggplot2)
## extract variance explained attribute
varexp <- attr(pc_analysis, 'input_varexp')
## extract PC data
pcs = pc_analysis[, c("pc1","pc2")]
## generate plot
pcs |> ggplot(aes(x = pc1, y = pc2)) +
  geom_point(size = 2, col = "#377EB8") +
  labs(x = paste0("PC1; VarExp =  ", varexp[1]*100, "%"), 
       y = paste0("PC2; VarExp =  ", varexp[2]*100, "%")) +
  theme_bw()
```



### Additional pc_and_outliers() attributes

In addition, the variance explained vector is appended to the returned `data.frame` as and `attribute`. This can be accessed with the attribute name: `[source_layer]_varexp`, in this case we used the `input` data, therefore the attribute name is `input_varexp`. In a similar way, the results of the acceleration analysis (`input_num_pcs_scree`) and a parallel analysis (`input_num_pcs_parallel`) can also be extracted.


```{r ScreePlot, out.width="100%", fig.align='center', warning=FALSE, message=FALSE, fig.alt = "Variance explained"}
library(ggplot2)

# extract varexp from attributes
varexp <- attr(pc_analysis, 'input_varexp')

# subset to top 100 for nicer plotting
if (length(varexp) > 100) varexp <- varexp[1:100]

# get acceleration and parallel analysis results
af <- attr(pc_analysis, 'input_num_pcs_scree')

# as data.frame
x_labs <- sub("(?i)pc","", names(varexp))
ve     <- data.frame("pc"      = factor(x_labs, levels=x_labs),
                     "var_exp" = varexp)
lines  <- data.frame("Analysis" = c("Acceleration"), 
                     "pc"       = c(af))   

# plot
ggplot(ve, aes(x = pc, y = var_exp)) +
  geom_line(color = "grey") +
  geom_point(shape = 21, fill = "#377EB8", size = 2) +
  geom_vline(data = lines, aes(xintercept = pc, color = Analysis), inherit.aes = FALSE) +
  scale_color_manual(values = c("Acceleration"="#E41A1C")) +
  scale_x_discrete(labels = function(x) ifelse(seq_along(x) %% 10 == 0 | x==1, x, "")) +
  labs(x = "PC", y = "Variance explained") +
  theme_classic() +
  theme(legend.position = "top")
```


## Run sample & feature summaries together

```{r sample_feature_summary}
sf_sum <- summarise(omiprep         = mydata, 
                    source_layer    = "input", 
                    outlier_udist   = 1.0,
                    tree_cut_height = 0.5,
                    sample_ids      = sids, ## It is also possible to run on a subset of samples and/or features
                    feature_ids     = NULL,
                    output          = "data.frame", 
                    cores           = 1)

## two data frames are returned as a list object
names(sf_sum)
```

### Table of sample summary on subset

Note that when the summarise() function is used the sample summary now includes PCA derived summary data. This is not the case when the sample_summary() function is run alone, as seen above. The reason for the difference is because the PCA data is dependent upon the feature_summary() analysis.

Also, please note that when running on a subset, you are returned the full summary for all samples and features, but only the summary data for the specified subset will be populated, the rest will be `NA`.

```{r sample_sum_subset_table2, echo=FALSE}
# note that for illustrative purposes we are using a log outlier unit distance of 1.0 here, in practice we tend to favor a value of 5.0.
sf_sum$sample_summary |> 
  head(n = 10) |> 
  knitr::kable( digits = 3, row.names = FALSE, align = "c") |>
  kableExtra::kable_styling(full_width = TRUE) 
```

